The \dealii{} Library, Version 9.3

### 2021-05-31 9.3/paper.tex
Further changes

A user-focused summary of information around simplex and mixed mesh
support is also available on the new module page at
  \url{https://www.dealii.org/current/doxygen/deal.II/group__simplex.html}
In particular, it shows how to solve a simple Poisson problem like in
the \texttt{step-3} tutorial program
on simplex and mixed meshes, with a focus on the necessary changes to
the workflow.
At the time of this release, there are also 92 tests (in the folder \texttt{tests/simplex})
targeting the new simplex and mixed mesh support. In particular, the folder also
contains ported variants of a number of existing tutorials: 1, 2, 3, 4, 6, 7, 8, 12, 17, 18, 20, 23, 31, 38,
40, 55, 67, 68, and 74.

@@

@@

    \foreach \i in {6,7,8,9,10,11,12,13,14}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

@@

  \caption{\it Reference cells in 2D (triangle, quadrilateral) and 3D
    (tetrahedron, pyramid, wedge, hexahedron) with the support points
    indicated both for linear ({\color{blue}$\bullet$}) and quadratic
    ({\color{red}$\times$}) shape functions.
  \todo[inline]{The pyramid is missing red crosses. The wedge is
    missing them on three edges.}}
  \label{fig:simplex}
\end{figure}

@@

FEValues<dim, spacedim> fe_values(mapping, fe, quad, flags);
\end{c++}
The  list of currently supported finite-element classes is provided in Table~\ref{tab:simplex:fe}. Currently,
only linear iso-parametric mappings (via the \texttt{MappingFE} and
\texttt{MappingFEField} classes) are available.
\todo[inline]{I don't understand the previous sentence. "linear" means
  p=1, but isoparametric means that it has the same p as the finite
  element. It can't be both.}
For quadrature, the classes \texttt{QDuffy}, \texttt{QGaussSimplex}, \texttt{QWitherdenVincentSimplex},
\texttt{QGaussPyamid}, and \texttt{QGaussWedge} are
available.

@@

The introduction of simplex and mixed meshes leads to some implications
for the user if these features are to be used. For instance, each cell might have a different type with
different number of vertices, lines, and faces so that these quantities can not be
compile-time constants anymore. This information used to be queried from
the \texttt{GeometryInfo} class. To avoid using this class, we have extended
relevant classes, e.g., \texttt{TriaAccessor} or \texttt{TriaCellAccessor},
with useful new functions like \texttt{n\_vertices()}, \texttt{n\_lines()}, or
\texttt{n\_faces()} so that users can simply write:
\begin{c++}
for(const auto & cell : tria.active_cell_iterators())
  for(unsigned int f = 0; f < cell->n_faces(); ++f)
    // do something with cell->face(f);
\end{c++}
Alternatively, one can use an iterator-based approach to loop over all faces
of a cell, introduced in the previous release. The relevant functions have been adjusted to be able to deal with
the now variable number of faces per cell:
\begin{c++}
for(const auto & cell : tria.active_cell_iterators())
  for(const auto & face : cell->face_iterators())
    // do something with face
\end{c++}
Furthermore, for mixed meshes, the number of degrees of freedom will differ between cells so that
cell-local arrays need to be resized for each
cell (as has previously already been the case in the $hp$-context):
\begin{c++}
std::vector<double> local_rhs;
for(const auto & cell : tria.active_cell_iterators())
 {
   hp_fe_values.reinit(cell);
   local_rhs.resize(cell->get_fe().dofs_per_cell);
 }
\end{c++}

@@

In contrast, the current release now also supports ``global
coarsening'' algorithms \citep{becker2000multigrid, sundar2012parallel} when
using continuous (\texttt{FE\_Q}, \texttt{FE\_SimplexP}) and
discontinuous (\texttt{FE\_DGQ}, \texttt{FE\_SimplexDGP})
elements. Global coarsening builds multigrid levels for the entire
domain, coarsening \textit{all} cells of a triangulation regardless of
how many times they have been refined. In addition, the framework now available in
\dealii{} in not only applicable to geometric coarsening, but can also
perform coarsening by reducing the polynomial degree ($p$-multigrid,
see~\cite{Fehn2020}), for example
to support $hp$-adaptive meshes.
%where it has been shown that one can
%efficiently precondition high-order spaces with lower-order finite
%element operators.
% MK: I do not think "precondition" is a precise enough description of p-multigrid
% because one still needs smoothing on the fine levels in addition to the
% "solve" at the coarser degree, a fact not transported by the mere
% 'precondition' term; rather than trying to discuss this in a more precise way,
% let us simply skip it as that is not central to the presentation here.
Finally, the implementation also supports transfer between continuous
and discontinuous elements as a further way to create multigrid levels.

@@

In the current release, specializations of these classes that do
not rely on the template parameters $k$ and $q$ (expressed in the code
using special values ``-1'' and ``0'') have been added. For example:
\begin{c++}
FEEvaluation<dim, -1, 0, n_components, Number, VectorizedArrayType>
  phi(range, dofhandler_index, quadrature_index, first_selected_component);
\end{c++}
These classes select at runtime -- for common low and medium
polynomial degree/quadrature combinations ($k\le 6$ and $q\in\{ k+1,
k+2, \left\lfloor (3k)/2 \right\rfloor \}$) -- efficient precompiled implementations and default to non-templated
evaluation kernels else.

@@

DEAL_II_NAMESPACE_OPEN
template struct dealii::internal::FEEvaluationFactory
  <dim, VectorizedArrayType::value_type, VectorizedArrayType>;
DEAL_II_NAMESPACE_CLOSE
\end{c++}

@@

Until now, the \texttt{FEEvaluation} classes used the template
parameters $k$ and $q$ to select the correct active FE and quadrature
index and cumbersome detection of subranges of the same degree.
\todo{I don't understand what ``detection of subranges of the same
  degree'' refers to. Please clarify. In particular, I don't know what
``subranges'' mean in this paragraph.}
The creation of subranges is now performed internally, and the non-templated versions
of the \texttt{FEEvaluation} classes have been extended for the $hp$-case. To determine
the desired FE and quadrature index of a subrange, the current cell/face range has to be provided
to the constructors of the \texttt{FEEvaluation} classes. These changes enable
users to write matrix-free code independently of whether $hp$-capabilities are used or not.

@@

To give an example of the new features, let us consider the evaluation of a
surface tension force in the context of sharp-interface methods, whose
contribution is added to a fluid solver by multiplication with test function
and addition over quadrature points located at the interface section
$\Gamma_K=\Gamma \cup K$ located on the current cell $K$ and that is,
in general, located differently within $K$ than for any other cell:
\begin{align*}
\left(\mathbf{v}, \kappa \, \mathbf{n}\right)_\Gamma
\approx
\sum_q \mathbf{v}(\mathbf{x}_q) \cdot \left(\kappa(\mathbf{x}_q) \, \mathbf{n}(\mathbf{x}_q)\right) (JxW)_q.
\end{align*}
In \dealii{}, this can now be conveniently written as
\begin{c++}
phi_normal.reinit(cell, reference_points);
phi_force.reinit(cell, reference_points);
phi_normal.evaluate(normal_values, EvaluationFlags::values);
for (unsigned int q = 0; q < n_points; ++q)
   phi_force.submit_value(phi_curvature.get_value(q) *
                          phi_normal.get_value(q) * JxW[q], q);
phi_force.integrate(force_values, EvaluationFlags::values);
\end{c++}
The quadrature points (at reference positions
\texttt{reference\_points}) and the related \texttt{JxW} value can,
for example, come from
a mesh of lower dimension. Determining to which \texttt{cell} a quadrature
point belongs to on the background mesh, including the reference cell
coordinates \texttt{reference\_points}, is aided by functions like
\texttt{GridTools::find\_all\_active\_cells\_around\_point()}. While these
functions have been available in \dealii{} previously, their performance has
been considerably enhanced with the aforementioned more optimized code paths
for selected mappings.

While \texttt{FEPointEvaluation} assumes that evaluation points are already sorted according to 
the owning cells and thus can focus on cell-local operations, the new class
\texttt{RemotePointEvaluation} is responsible for determine the owning cells in a distributed
context and for providing efficient communication patterns for the data exchange. In \texttt{deal.II},
the class has been successfully applied together with \texttt{FEPointEvaluation} to evaluate a distributed
solution vector at arbitrary points (see also the
\texttt{VectorTools::evaluate\_at\_points()} function).

@@

\item \texttt{step-71} focuses on automatic and symbolic
  differentiation (AD and SD, in short) as a tool to make solvers for complex,
  nonlinear problems possible. To this end, \dealii{} can interface to a number
  of AD and SD libraries, specifically Trilinos' Sacado package
  \cite{Bartlett2006a}, ADOL-C \cite{Griewank1996a}, and
  SymEngine \cite{symengine-web-page}. The tutorial
  illustrates how these techniques can be used to compute derivatives
  of first a rather simple function, and then of the much more complex
  energy functions of two magnetoelastic and magneto-viscoelastic
  material formulations in which just the scalar energy functional
  takes up the better part of a page, and even first derivatives can
  only be computed with heroic effort.

>>>>

A user-focused summary of information around simplex and mixed mesh
support is also available on the new module page at
  \url{https://www.dealii.org/current/doxygen/deal.II/group__simplex.html}.
In particular, it shows how to solve a simple Poisson problem like in
the \texttt{step-3} tutorial program
on simplex and mixed meshes, with a focus on the necessary changes to
the workflow.
At the time of this release, there are also 92 tests (in the folder \texttt{tests/simplex})
targeting the new simplex and mixed mesh support. In particular, the folder also
contains ported variants of a number of existing tutorials: 1, 2, 3, 4, 6, 7, 8, 12, 17, 18, 20, 23, 31, 38,
40, 55, 67, 68, and 74.

@@

    \coordinate (15) at (0,0,0.75);
    \coordinate (16) at (1,0,0.75);
    \coordinate (17) at (0.0,1.0,0.75);

@@

    \foreach \i in {6,7,8,9,10,11,12,13,14,15,16,17}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

@@

  \caption{\it Reference cells in 2D (triangle, quadrilateral) and 3D
    (tetrahedron, pyramid, wedge, hexahedron) with the support points
    indicated both for linear ({\color{blue}$\bullet$}) and quadratic
    ({\color{red}$\times$}) shape functions.
  %\todo[inline]{The pyramid is missing red crosses. The wedge is
  %  missing them on three edges.}
  % PM: we only have linear pyramids for now.
  }
  \label{fig:simplex}
\end{figure}

@@

FEValues<dim, spacedim> fe_values(mapping, fe, quad, flags);
\end{c++}
The  list of currently supported finite-element classes is provided in Table~\ref{tab:simplex:fe}. Currently,
only linear and quadratic mappings via the \texttt{MappingFE} and
\texttt{MappingFEField} classes built around the listed nodal elements are available.
%\todo[inline]{I don't understand the previous sentence. "linear" means
%  p=1, but isoparametric means that it has the same p as the finite
%  element. It can't be both.}
For quadrature, the classes \texttt{QDuffy}, \texttt{QGaussSimplex}, \texttt{QWitherdenVincentSimplex} \cite{schloemer21, witherden2015identification},
\texttt{QGaussPyamid}, and \texttt{QGaussWedge} are
available.

@@

The introduction of simplex and mixed meshes leads to some implications
for the user if these features are to be used. For instance, each cell might have a different type with
different number of vertices, lines, and faces so that these quantities can not be
compile-time constants anymore. This information used to be queried from
the \texttt{GeometryInfo} class. To avoid using this class, we have extended
relevant classes, e.g., \texttt{TriaAccessor} or \texttt{TriaCellAccessor},
with useful new functions like \texttt{n\_vertices()}, \texttt{n\_lines()}, or
\texttt{n\_faces()} so that users can simply write:
\begin{c++}
for(const auto & cell : tria.active_cell_iterators())
  for(unsigned int f = 0; f < cell->n_faces(); ++f)
    // do something with cell->face(f);
\end{c++}
Alternatively, one can use an iterator-based approach to loop over all faces
of a cell, introduced in the previous release. The relevant functions have been adjusted to be able to deal with
the now variable number of faces per cell:
\begin{c++}
for(const auto & cell : tria.active_cell_iterators())
  for(const auto & face : cell->face_iterators())
    // do something with face
\end{c++}
Furthermore, for mixed meshes, the number of degrees of freedom will differ between cells so that
cell-local arrays need to be resized for each
cell (as has previously already been the case in the $hp$-context):
\begin{c++}
std::vector<double> local_rhs;
for(const auto & cell : tria.active_cell_iterators())
 {
   hp_fe_values.reinit(cell);
   local_rhs.resize(cell->get_fe().n_dofs_per_cell());
 }
\end{c++}

@@

In contrast, the current release now also supports ``global
coarsening'' algorithms \citep{becker2000multigrid, sundar2012parallel} when
using continuous (\texttt{FE\_Q}, \texttt{FE\_SimplexP}) and
discontinuous (\texttt{FE\_DGQ}, \texttt{FE\_SimplexDGP})
elements. Global coarsening builds multigrid levels for the entire
domain, coarsening \textit{all} cells of a triangulation regardless of
how many times they have been refined. In addition, the framework now available in
\dealii{} is not only applicable to geometric coarsening, but can also
perform coarsening by reducing the polynomial degree ($p$-multigrid,
see~\cite{Fehn2020}), for example
to support $hp$-adaptive meshes.
%where it has been shown that one can
%efficiently precondition high-order spaces with lower-order finite
%element operators.
% MK: I do not think "precondition" is a precise enough description of p-multigrid
% because one still needs smoothing on the fine levels in addition to the
% "solve" at the coarser degree, a fact not transported by the mere
% 'precondition' term; rather than trying to discuss this in a more precise way,
% let us simply skip it as that is not central to the presentation here.
Finally, the implementation also supports transfer between continuous
and discontinuous elements as a further way to create multigrid levels.

@@

In the current release, specializations of these classes that do
not rely on the template parameters $k$ and $q$ (expressed in the code
using special values ``-1'' and ``0'') have been added. For example:
\begin{c++}
FEEvaluation<dim, -1, 0, n_components, Number, VectorizedArrayType>
  phi(range, dofhandler_index, quadrature_index, first_selected_component);
\end{c++}
These classes select at runtime -- for common low and medium
polynomial degree/quadrature combinations ($k\le 6$ and $q\in\{ k+1,
k+2, \left\lfloor (3k)/2 \right\rfloor \}$) -- efficient precompiled implementations and default to non-templated
evaluation kernels else (see also \texttt{FEEvaluation::fast\_evaluation\_\allowbreak supported()}).

@@

DEAL_II_NAMESPACE_OPEN
template struct dealii::internal::FEEvaluationFactory
  <dim, VectorizedArrayType::value_type, VectorizedArrayType>;
// same for FEFaceEvaluationFactory and CellwiseInverseMassFactory (skipped)
DEAL_II_NAMESPACE_CLOSE
\end{c++}

@@

Until now, the \texttt{FEEvaluation} classes used the template
parameters $k$ and $q$ to select the correct active FE and quadrature
index and users were responsible for the cumbersome detection of 
subranges of the same $k$ and $q$ within the cell ranges returned by the 
matrix-free loops.
%\todo{I don't understand what ``detection of subranges of the same
%  degree'' refers to. Please clarify. In particular, I don't know what
%``subranges'' mean in this paragraph.}
The creation of subranges is now performed internally, and the non-templated versions
of the \texttt{FEEvaluation} classes have been extended for the $hp$-case. To determine
the desired FE and quadrature index of a subrange, the current cell/face range has to be provided
to the constructors of the \texttt{FEEvaluation} classes. These changes enable
users to write matrix-free code independently of whether $hp$-capabilities are used or not.

@@

To give an example of the new features, let us consider the evaluation of a
surface tension force in the context of sharp-interface methods, whose
contribution is added to a fluid solver by multiplication with test function
and addition over quadrature points located at the interface section
$\Gamma_K=\Gamma \cup K$ located on the current cell $K$ and that is,
in general, located differently within $K$ than for any other cell:
\begin{align*}
\left(\mathbf{v}, \kappa \, \mathbf{n}\right)_\Gamma
\approx
\sum_q \mathbf{v}(\mathbf{x}_q) \cdot \left(\kappa(\mathbf{x}_q) \, \mathbf{n}(\mathbf{x}_q)\right) (JxW)_q.
\end{align*}
In \dealii{}, this can now be conveniently written as
\begin{c++}
phi_normal.reinit(cell, reference_points);
phi_force.reinit(cell, reference_points);
phi_normal.evaluate(normal_values, EvaluationFlags::values);
for (unsigned int q = 0; q < n_points; ++q)
   phi_force.submit_value(phi_curvature.get_value(q) *
                          phi_normal.get_value(q) * JxW[q], q);
phi_force.integrate(force_values, EvaluationFlags::values);
\end{c++}
The quadrature points (at reference positions
\texttt{reference\_points}) and the related \texttt{JxW} value can,
for example, come from
a mesh of lower dimension. Determining to which \texttt{cell} a quadrature
point belongs to on the background mesh, including the reference cell
coordinates \texttt{reference\_\allowbreak points}, is aided by functions like
\texttt{find\_active\_cell\_around\_point()}
\texttt{find\_all\_active\_cells\_\allowbreak around\_\allowbreak point()} from the \texttt{GridTools} namespace. While these
functions have been available in \dealii{} previously, their performance has
been considerably enhanced with the aforementioned more optimized code paths
for selected mappings.

While \texttt{FEPointEvaluation} assumes that evaluation points are already sorted according to 
the owning cells and thus can focus on cell-local operations, the new class
\texttt{RemotePointEvaluation} is responsible for determine the owning cells in a distributed
context and for providing efficient communication patterns for the data exchange. In \texttt{deal.II},
the class has been successfully applied together with \texttt{FEPointEvaluation} to evaluate a distributed
solution vector at arbitrary points (see also the
\texttt{VectorTools::point\_values()} function).

@@

\item \texttt{step-71} focuses on automatic and symbolic
  differentiation (AD and SD, in short) as a tool to make solvers for complex,
  nonlinear problems possible. To this end, \dealii{} can interface to a number
  of AD and SD libraries, specifically Trilinos' Sacado package
  \cite{Bartlett2006a}, ADOL-C \cite{Griewank1996a}, and
  SymEngine \cite{symengine-web-page}. The tutorial
  illustrates how these techniques can be used to compute derivatives
  first of a rather simple function, and then of the much more complex
  energy functions of two magnetoelastic and magneto-viscoelastic
  material formulations in which just the scalar energy functional
  takes up the better part of a page, and even first derivatives can
  only be computed with heroic effort.
